Bridging the Gap Between Visual and Analytical Machine Learning Testing

### 2023-10-25 report/icst24.tex
feat(report): 2023-10-25 17:10 updates

Testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect $54,070$ Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occuring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part. Our results indicate the need for automated tools that can reduce the manual effort required to verify visualisations used to test ML systems. And help ML practitioners formulate better analytical assertions.

@@

% TODO still needs improvement
This study uses Jupyter Notebooks mined from Github. Jupyter notebooks have been widely adopted by the Data Science and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}. Computational notebooks provide a rich source of data in the form of natural text, code and visualisations. Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}. We leverage the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.

@@

We programmatically parse the JSON structure of the notebooks (more details regarding the internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}) and represent each notebook as a pandas dataframe for further analysis. Each cell is mapped as a row in the dataframe. For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe. Notebooks without any visualisations are excluded by checking if the image column of  the dataframe is empty.

@@

We malysed the source code of the visualisation and assertion code cells. And excluded VA pairs that were not related to each other. We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls. In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation. Listing~\ref{lst:prompt} shows the prompt used for such cases. In a few cases, where the explaination of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair. Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.

@@

% TODO add the 54K NB found with assert here as well

From the 1764 Jupyter notebooks that were analysed in this study, we found 269 VA pairs that were semantically related to each other.

@@

\begin{table}
  \centering
  \caption{Most frequently occurring assertions that use a comparison operator.}
  \begin{tabular}{@{}m{0.9\linewidth}@{}}
    \toprule
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert accuracy_score(y_test, pred) > 0.95\end{lstlisting}\\
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert HISTORY['test_accuracy'][-1] > 0.92\end{lstlisting}\\
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert np.mean(mean_rw_history[-10:]) > 10.
print("That's good enough for tutorial.")\end{lstlisting}\\
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."\end{lstlisting}\\
    \begin{lstlisting}[language=python,belowskip=0pt,aboveskip=0pt]
assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."\end{lstlisting}\\
    \bottomrule
  \end{tabular}
  \label{tab:compare-op-asserts}
\end{table}

@@

An overwhelming majority of assertions compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author\luis{we need to absolute values here (out of 269)}. This covers 46\% of the assertions with a comparison operator. It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook. The threshold could have been derived based on domain expertise. Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{CITEME}.\luis{Agathe Balayn might have something on these lines}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
  \caption{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
\end{figure}

Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window. The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set. A linear relationship between the ground truth and predictions is observed, indicating that the model was able to learn certain patterns from the training data. The spread of $y_{pred}$ at any given $y_{true}$ is wide. For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120. This Indicates that the model was not able to fit the training data that well, and can be improved with fine-tuning.

@@

The results from RQ2 provide empirical evidence to support the use of visualisations to test various properties of ML systems.  Visualisations are an efficient and effective medium to present complex information, that is easier to interpret by humans with the aid of visual cues. Visualisations used to test ML system properties however, demand manual verification effort. Information gained from visualisations and subsequent decisions made based on this information, must be manually verified in every iteration of the ML development cycle.

@@

In the 1764 notebooks analysed in this study, several visualisations lack assertions. However, unformalised expectations still exist within each visualisation which cannot be readily validated by other participants. ML testing demands proficiency in Machine Learning, Software Testing, and the underlying Mathematical concepts. This highlights the need for refining the education for these professionals and crafting tools to compensate for potential gaps in their knowledge.

@@

% TODO \luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}

@@

For example, a linear correlation between the true and predicted labels is evident in Figure~{fig:r2}. Yet, its formal assertion in Listing~\ref{lst:r2} necessitates defining an appropriate threshold for the coefficient of determination metric. Our results show that current software testing practices fail to capture the idiosyncrasies of ML code. There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.

@@

The empirical analysis presented in this study uses Jupyter notebooks mined from Github. To the best of our knowledge,~\cite{pimentel2019large} and~\cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks. We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github. However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}). We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targetted towards the ML and DS community.

In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword. Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}). Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in several Python testing libraries.

@@

This study study revealed the emerging practice of crafting analytical assertions based on visual insights to bridge the gap between qualitative and quantitative aspects of ML testing.

We identified 269 visualisation-assertion pairs from 1.7K Jupyter notebooks, that were semantically related to one another. And identified three recurring testing patterns within these pairs. Nearly 40\% of the assertions used ``magic'' thresholds, 75\% of the assertions used comparison operators, and 16\% of the assertions used external testing libraries.

This study serves as a first step in comprehending the landscape of ML testing in Notebooks, emphasizing the potential for the development of more effective tools and practices to streamline the validation of visualisations and elevate the quality of analytical assertions.

>>>>

Testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect $54,070$ Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part. Our results indicate the need for automated tools that can reduce the manual effort required to verify visualisations used to test ML systems. And help ML practitioners formulate better analytical assertions.

@@

Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}. Computational notebooks provide a rich source of data in the form of natural text, code and visualisations. Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}. This study leverages the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.

@@

We programmatically parse the JSON structure of the notebooks and represent each notebook as a pandas dataframe for further analysis. Each cell is mapped as a row in the dataframe. For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe. Notebooks without any visualisations are excluded by checking if the image column of  the dataframe is empty.

@@

We analysed the source code of the visualisation and assertion code cells. And excluded VA pairs that were not related to each other. We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls. In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation. Listing~\ref{lst:prompt} shows the prompt used for such cases. In a few cases, where the explanation of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair. Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.

@@

Although Github contains more than a million Jupyter notebooks, only 54K notebooks contained the keyword ``assert'' in them. From the 1764 Jupyter notebooks that were analysed in this study, we found 269 VA pairs that were semantically related to each other.

@@

\begin{table}
  \centering
  \caption{Most frequently occurring assertions that use a comparison operator.}
  \begin{tabular}{@{}m{0.9\linewidth}@{}}
    \toprule
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert accuracy_score(y_test, pred) > 0.95\end{lstlisting}\\
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert HISTORY['test_accuracy'][-1] > 0.92\end{lstlisting}\\
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert np.mean(mean_rw_history[-10:]) > 10.
print("That's good enough for tutorial.")\end{lstlisting}\\
    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."\end{lstlisting}\\
    \begin{lstlisting}[language=python,belowskip=0pt,aboveskip=0pt]
assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
    \end{lstlisting}\\
    \bottomrule
  \end{tabular}
  \label{tab:compare-op-asserts}
\end{table}

@@

An overwhelming majority of assertions compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author\luis{we need to absolute values here (out of 269)}. This covers 46\% of the assertions with a comparison operator. It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook. The threshold could have been derived based on domain expertise. Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{vogelsang2019requirementsi,rahimi2019toward}.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
  \caption{Scatter plot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
\end{figure}

Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window. The visualisation shows a scatter plot of the real and predicted values of $NO_2$ from the test set. A linear relationship between the ground truth and predictions is observed, indicating that the model was able to learn certain patterns from the training data. The spread of $y_{pred}$ at any given $y_{true}$ is wide. For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120. This Indicates that the model was not able to fit the training data that well, and can be improved with fine-tuning.

@@

The results from RQ1 provide empirical evidence to support the use of visualisations to test various properties of ML systems. Visualisations are an efficient and effective medium to present complex information, that is easier to interpret by humans with the aid of visual cues. Visualisations used to test ML system properties however, demand manual verification effort. Information gained from visualisations and subsequent decisions made based on this information, must be manually verified in every iteration of the ML development cycle.

@@

In the 1764 notebooks analysed in this study, several visualisations lack assertions. However, implicit expectations still exist within each visualisation which cannot be readily validated by other participants. ML testing demands proficiency in Machine Learning, Software Testing, and the underlying Mathematical concepts. This highlights the need for refining the education for these professionals and crafting tools to compensate for potential gaps in their knowledge.

@@

@@

For example, a linear correlation between the true and predicted labels is evident in Figure~\ref{fig:r2}. Yet, its formal assertion in Listing~\ref{lst:r2} necessitates defining an appropriate threshold for the coefficient of determination metric. Our results show that current software testing practices fail to capture the idiosyncrasies of ML code. There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.

@@

The empirical analysis presented in this study uses Jupyter notebooks mined from Github. To the best of our knowledge,~\cite{pimentel2019large} and~\cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks. We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github. However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}). We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targeted towards the ML and DS community.

In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword. Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}). Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an exhaustive list of keywords that appear in several Python testing libraries.

@@

This study empirically supports the efficacy of visualisations for testing various properties of machine learning systems. Visualisations enhance the ability of ML practitioners to interpret complex information through visual cues. However, decisions based on such visualisations demand manual verification in each development iteration. Analytical assertions derived from ML visualisations can reduce this manual effort by encapsulating key insights, and ensuring continuity in interpretation across different development stages and project collaborators.

Our results indicate that crafting analytical assertions from ML visualisations, is an emerging testing technique. Majority of the assertions however, are found to be primitive and unable to convey all information discerned from their visual counter-parts. This study finds a gap in the current software testing methods to address the unique challenges of ML. And emphasises the urgent need for automated tools to bridge visual assessments and analytical assertions.
